Coursera Reproducible Research Peer Assessment Assignment 1
========================================================
Part 1 - Read in and process the data
-------------------------------------

```{r}
#read in the data and produce a quick summary
activityData <- read.csv("activity.csv", colClasses = c("numeric", "Date", "numeric"), header = T)
summary(activityData)
```

Part 2 - What is the mean total number of steps taken each day?  
---------------------------------------------------------------
The total number of steps taken per are calculated, and a plot of these totals produced.

```{r fig.height = 4, fig.width = 6}
#produce an array containing daily sums, and make a histogram
daySum <- tapply(activityData$steps, activityData$date, sum)
hist(daySum, main= "Daily Steps", xlab = "Total number of Daily Steps", col = "blue")
```

The mean and median are determined, ignoring missing values.

```{r}
#calculate the mean
mean(daySum, na.rm = T)
#calculate the median
median(daySum, na.rm = T)
```

Part 3 - What is the average daily activity pattern?  
----------------------------------------------------
This can be answered by first looking at a time series plot of the average number of steps taken in each 5-minute time interval, again ignoring missing values.

```{r fig.height = 4, fig.width = 6}
#construct a data frame of the average steps for each interval
library(plyr)
intMean <- ddply(activityData, .(interval), summarise, mean = mean(steps, na.rm = T))
#plot the steps vs. interval
plot(intMean$interval, intMean$mean, type = "l", main = "Average Steps per Interval", xlab = "Interval (min since 00:00)", ylab = "Average Number of Steps")
```

The 5-minute interval in which the maximum number of steps were taken:

```{r}
#determine the interval with the maximum average number of steps
intMean[(intMean$mean == max(intMean$mean)),]
```

Part 4 - Inputing missing values.  
---------------------------------
Several of the intervals,usually by entire days, are missing values for the number of steps.
The total number of missing values for each variable:

```{r}
#sum the NA's in each column variable
colSums(is.na(activityData))
```

A simple method with which to replace the missing values is to use the mean value for that time interval. Once a new data frame is made with the values are replaced,

```{r}
#create a new data frame with the missing values replaced
aData2 <- activityData
aData2$steps <- replace(activityData$steps, is.na(activityData$steps), round(intMean$mean,1))
```

The previous look at the total daily steps can be reapplied.

```{r fig.width = 6, fig.height = 4}
hist(tapply(aData2$steps, aData2$date, sum),main= "Daily Steps with Missing Values Replaced", xlab = "Total number of Daily Steps", col = "red")
```

And the mean and median values recalculated:

```{r}
#calculate the mean
mean(tapply(aData2$steps, aData2$date, sum))
#calculate the median
median(tapply(aData2$steps, aData2$date, sum))
```

And we can see that there appears to very little bias introduced by the missing values, at least when compared to a new data set with a very simple replacement scheme, and using only the simple measures of mean and median for this comparison.

Part 5 - Are there differences in activity patterns between weekdays and weekends?  
----------------------------------------------------------------------------------
This is examined through creation of a set of two factor variables, weekday and weekend, and assigning the data to each one, based on the type of day. A simple panel plot demonstrates clear differences between the patterns of the average number of steps in each interval, averaged across the weekedays, compared to the weekend.

```{r}
#create factors based on day of the week, and assign days
aData2$dayType <- ifelse(weekdays(aData2$date) == "Sunday" | weekdays(aData2$date) == "Saturday", "weekend", "weekday")
aData2$dayType <- factor(aData2$dayType)
dayTypeMean <- ddply(aData2, .(interval, dayType), summarise, mean = mean(steps))

#plot steps based on the day factor
library(lattice)
xyplot(dayTypeMean$mean ~ dayTypeMean$interval | dayTypeMean$dayType, type = "l", xlab = "Interval (min since 00:00)", ylab = "Average Number of Steps")
```